---
title: "Titanic Tragedy Dataset EDA and Training"
output: html_notebook
      ---
# LOADING DATASET
```{r}
titanic <- read.csv("/home/tanmay/Datasets/train.csv")
```

# EXPLORATORY DATA ANALYSIS

```{r}
#DATASET PREVIEW
head(titanic)
```
## DATASET STRUCTURE
```{r}
str(titanic)
```
## DATASET SUMMARY
```{r}
summary(titanic)
```
## CHECKING MISSING VALUES
```{r}
checkNA <- function(x){sum(is.na(x))/length(x)*100} 
sapply(titanic,checkNA)
print("MISSING VALUES WITHOUT NA")
checkMissing <- function(x){sum(x=="")/length(x)*100}
sapply(titanic,checkMissing)
```
## Missing Value Treatment
```{r}
#1. Age: Replacing NA values in Age with mean 
#titanic[is.na(titanic$Age),6] <- mean(titanic$Age)
titanic$Age[is.na(titanic$Age)] <- round(mean(titanic$Age, na.rm = TRUE))

#2. Embarked: Replacing Empty Embarked with most common value 'S'
titanic$Embarked <- replace(titanic$Embarked, which(titanic$Embarked==""), 'S')
```

```{r}
Title <-  gsub("^.*, (.*?)\\..*$", "\\1", titanic$Name)
titanic$Title <- as.factor(Title)
table(Title)
```

```{r}
  titanic$FamilyCount <-titanic$SibSp + titanic$Parch + 1 
titanic$FamilySize[titanic$FamilyCount == 1] <- 'Single' 
titanic$FamilySize[titanic$FamilyCount < 5 & titanic$FamilyCount >= 2] <- 'Small' 
titanic$FamilySize[titanic$FamilyCount >= 5] <- 'Big' 
titanic$FamilySize=as.factor(titanic$FamilySize)
table(titanic$FamilySize)
```

## DATA PREPROCESSING
```{r}
# 1.Changing names of few categorical variables for interpretability 
titanic$Survived <- ifelse(titanic$Survived==1,"Yes","No")
titanic$Survived <- as.factor(titanic$Survived)

titanic$Embarked <- ifelse(titanic$Embarked=="S","Southampton", 
                              ifelse(titanic$Embarked=="C","Cherbourg", "Queenstown"))
titanic$Embarked <- as.factor(titanic$Embarked)
# 
# # 2.Converting categorical variables from int to factor
# i) Pclass
titanic$Pclass <- as.factor(titanic$Pclass)
# 
# # ii) SibSp
titanic$SibSp <- as.factor(titanic$SibSp)
# 
# # iii) Parch
titanic$Parch <- as.factor(titanic$Parch)

#Preview
head(titanic)
```
  
## Survival Demographic

```{r}

library(ggplot2)
ggplot(titanic, aes(Survived,fill = Survived))+
  geom_bar()+
  ggtitle("Barplot to represent Passenger Count who Survived vs who Died")
```
## Average Age of Passengers
```{r}
summary(titanic$Age)
d <- density(titanic$Age)
plot(d,main="Passenger Age Distribution",xlab="Age",ylab="Frequency",col="blue")
```

## Proportion of survivors by gender
```{r}
ggplot(titanic, aes(x=Sex,fill=Survived))+ geom_bar(position = "dodge") + geom_text(stat='count',aes(label=..count..),position = position_dodge(0.9),vjust=-0.2) +
ylab("Number of Passengers")
```
## Age distribution of Survivors & Non-Survivors
```{r}
ggplot(titanic) + geom_freqpoly(mapping = aes(x = Age, color = Survived), binwidth = 2.5) +
ylab("Frequency")
```

## Distribution of Passenger Fare for Survivors & Non-Survivors
```{r}
ggplot(titanic) + geom_freqpoly(mapping = aes(x = Fare, color = Survived), binwidth = 10)
```

## Passenger Class of most Non-Survivors
```{r}
  ggplot(titanic, aes(x=Pclass,fill=Survived))+ geom_bar(position = "dodge") + geom_text(stat='count',aes(label=..count..),position = position_dodge(0.9),vjust=-0.2) +
ylab("Number of Passengers") + xlab("Passenger Class")
```

## Proportion of survivors by place of Embarkment
```{r}
ggplot(titanic, aes(x=Embarked,fill=Survived))+ geom_bar(position = "dodge") + geom_text(stat='count',aes(label=..count..),position = position_dodge(0.9),vjust=-0.2) +
ylab("Number of Passengers")
```
## Number of sibling/spouses aboard Titanic & Passenger Class

```{r}
ggplot(titanic, aes(x=SibSp,fill=Pclass))+ geom_bar(position = "dodge") + geom_text(stat='count',aes(label=..count..),position = position_dodge(0.9),vjust=-0.2) +
ylab("Number of Passengers") + xlab("Number of Siblings")
```
  
## Number of sibling/spouses aboard Titanic related to Survival
```{r}
ggplot(titanic, aes(x=SibSp,fill=Survived))+ geom_bar(position = "dodge") + geom_text(stat='count',aes(label=..count..),position = position_dodge(0.9),vjust=-0.2) +
ylab("Number of Passengers")+xlab("Number of Siblings/Spouse")
```

## Number of parents/children aboard Titanic differ with Passenger Class
```{r}
ggplot(titanic, aes(x=Parch,fill=Pclass))+ geom_bar(position = "dodge") + geom_text(stat='count',aes(label=..count..),position = position_dodge(0.9),vjust=-0.2) +
ylab("Number of Passengers") + xlab("Number of Parents/Children")
```

## Relation between Passenger Name Title & Survival
```{r}
ggplot(titanic, aes(x=Title,fill=Survived))+ geom_bar(position = "dodge") + geom_text(stat='count',aes(label=..count..),position = position_dodge(0.9),vjust=-0.2) +
ylab("Number of Passengers") + xlab("Passenger Name Title") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Relation between Family Size & Survival
```{r}
ggplot(titanic, aes(x=FamilySize,fill=Survived))+ geom_bar(position = "dodge") + geom_text(stat='count',aes(label=..count..),position = position_dodge(0.9),vjust=-0.2) +
ylab("Number of Passengers") + xlab("Family Size")
```

# EDA COMPLETE
## ________________________________________________________________________________________________________________________________________________________________

# USING LOGISTIC REGRESSION FOR TARGET VARIABLE ("SURVIVED")

## CREATING DUMMIES OF CATEGORICAL COLUMNS

```{r}
library(dummies)
complete_data <- read.csv("/home/tanmay/Datasets/train.csv")

## Missing values imputation
complete_data$Embarked[complete_data$Embarked==""] <- "S"
complete_data$Age[is.na(complete_data$Age)] <- median(complete_data$Age,na.rm=T)

## Removing Cabin as it has very high missing values, passengerId, Ticket and Name are not required
library(dplyr)
titanic_data <- complete_data %>% select(-c(Cabin, PassengerId, Ticket, Name))

## Converting "Survived","Pclass","Sex","Embarked" to factors
for (i in c("Pclass","Sex","Embarked")){
  titanic_data[,i]=as.factor(titanic_data[,i])
}

## Create dummy variables for categorical variables
library(dummies)
titanic_data <- dummy.data.frame(titanic_data, names=c("Pclass","Sex","Embarked"), sep="_")
```

```{r}
head(titanic_data)
```

## SPLITTING DATA INTO TRAINING AND TESTING

```{r}
  training_data = titanic_data[1:790,]
  surv.x = training_data[,-1]
  surv.y = training_data[,1]
  testing_data = titanic_data[791:891,]
  test.x = testing_data[,-1]
  test.y = testing_data[,1]
```

## Building Helper Functions

```{r}
library(ggplot2)
library(dplyr)

#sigmoid function, inverse of logit
sigmoid <- function(z){1/(1+exp(-z))}

#cost function
cost <- function(theta, X, y){
  m <- length(y) # number of training examples
  h <- sigmoid(X %*% theta)
  J <- (t(-y)%*%log(h)-t(1-y)%*%log(1-h))/m
  J
}

#gradient function
grad <- function(theta, X, y){
  m <- length(y) 
  
  h <- sigmoid(X%*%theta)
  grad <- (t(X)%*%(h - y))/m
  grad
}
```


```{r}
# probability of getting 1
logisticProb <- function(theta, X){
  X <- na.omit(X)
  #add bias term and convert to matrix
  X <- mutate(X, bias =1)
  X <- as.matrix(X[,c(ncol(X), 1:(ncol(X)-1))])
  return(sigmoid(X%*%theta))
}

# y prediction
logisticPred <- function(prob){
  return(round(prob, 0))
}
```

## Logistic Regression Code

```{r}
logisticReg <- function(X, y){
  #remove NA rows
  X <- na.omit(X)
  y <- na.omit(y)
  #add bias term and convert to matrix
  X <- mutate(X, bias =1)
  #move the bias column to col1
  X <- as.matrix(X[, c(ncol(X), 1:(ncol(X)-1))])
  print(dim(X))
  # X <- as.matrix(X)
  y <- as.matrix(y)
  #initialize theta
  theta <- matrix(rep(0, ncol(X)), nrow = ncol(X))
  print(theta)
  #use the optim function to perform gradient descent
  costOpti <- optim(theta, fn = cost, gr = grad, X = X, y = y)
  #return coefficients
  return(costOpti$par)
}
```
## Feed Data to Logistic Reg Function

```{r}
mod <- logisticReg(surv.x, surv.y)
mod
```
```{r}
pre <- data.frame(matrix(c(1,0,0,0,1,24,0,0,60,1,0,0),nrow=1))
zpr <- logisticProb(mod,pre)
zpr
ans <- logisticPred(zpr)
ans
```

```{r}
  grid <- test.x
  prob <- logisticProb(mod,grid)
  print(length(prob))
  Z <- logisticPred(prob)
  print(length(Z))
  gridPred = cbind(grid, Z)
```
```{r}
results.table <- table(Z, test.y,dnn = c('Predicted','Actual'))
print(results.table)
```


```{r}
precision <- results.table[2,2] / (results.table[2,2] + results.table[2,1])
recal <- results.table[2,2] / (results.table[2,2] + results.table[1,2])
F1 <- 2 * precision * recal / (precision + recal)   
print(paste('F1-score: ', F1))

#My Accuracy Function
accu <- function(tp,tn,tot){
  return((tp+tn)/tot)
}
acc <- accu(tp = results.table[1,1],tn = results.table[2,2],length(test.y))
print(paste('Accuracy: ', acc))
```
```{r}
fourfoldplot(results.table, color = c("#CC6666", "#99CC99"),
             conf.level = 0, margin = 1, main = "Confusion Matrix")
```

